---
title: "Practical R"
author: "Abhijit Dasgupta"
date: "January-March, 2021"
output: 
  html_document:
    includes:
      after_body: footer.html
    theme: cosmo
    highlight: textmate

---

```{r, include = FALSE}
# Enables Font Awesome in HTML based bookdown outputs.
fa <- rmarkdown::html_dependency_font_awesome()
# fa$attachment <- "fonts/fontawesome-webfont.ttf"
htmltools::tagList(fa)
```

This is a six-week training program and tutorial in R delivered through the Data Science Academy at AstraZeneca. It will comprise a 2-hour teaching session including labs and examples on Mondays, and a 
follow-up office hours to go through exercises and labs on Friday. 

### Contact info: 

- Abhijit Dasgupta: via Microsoft Teams, e-mail at abhijit.dasgupta@astrazeneca.com


-------

<a href="http://www.r-project.org" target="_blank">R</a> is one of the most popular programming languages for 
data science and statistics. It is a free and open source language available for 
all major operating systems. It is extensible through a packaging system that has resulted
in over 12,000 user-contributed packages available on [CRAN](http://cran.r-project.org). 

In this course we will provide a practical approach to learning R using examples, with
the intention to get you using R for your data analyses, data visualization and reporting 
as quickly as possible. We will try to avoid the more esoteric and "computer science-y" aspects
of this programming language. We hope this approach and personal practice enables 
you to get started using R for your own work, find resources you may need beyond this
course to help you answer your particular questions, and develop workflows for your data
analyses that allow you to generate tables and figures for research papers, for example. 

The instructor uses R in his daily work for
many, many years and has a pretty detailed knowledge of R's advantages, pitfalls, tricks 
and shortcuts. Don't hesitate to reach out and ask your questions. 


------

```{r, fig.width=5, fig.height=4}
library(ggplot2)
data(iris)
ggplot(iris, aes(Sepal.Width, Sepal.Length, color=Species))+geom_point()+
  labs(x='Sepal Width', y = 'Sepal Length', color='Species')
```

-----

<p style="font-size: 80%"> This website is written using R, <a href="http://rmarkdown.rstudio.com/index.html" target="_blank">R Markdown</a> and various packages provided by <a href="http://www.rstudio.com" target="_blank">RStudio</a>. 
The source code is available on <a href="https://github.com/araastat/BIOF339/tree/gh-pages" target="_blank">GitHub</a> </p>
