---
title: "Course information"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(gt)
start_date = as_date('2021-01-25')
class_dates = start_date + (0:5)*days(7)

```


## Course description

The goal of this course is to introduce R as an analysis platform and tool for data science rather than purely a programming language. Throughout the course, emphasis will be placed on example-driven learning. Topics to be covered include: installation of R and R packages; R commands; R data types; loading data in R; manipulating data; exploring data through visualization; descriptive and inferential statistics;  building models; generating publication-quality graphics; creating reports using R Markdown. No prior programming experience is required

## Learning materials

There are no prescribed texts for this course. However we will draw on material from several texts 
and other materials described on the [Resources](resources.html) page, which are freely
available on the Internet. 

There are also several cheatsheets that can be used as resources, as well as online 
material that we will use for practice

### Course Goals

When you complete the course successfully, you will be able to:

+ Run R and RStudio, making use of inherent R features
+ Find and make use of the extensive packages (R add-ons) available for analyzin
g biological and other forms of data
+ Load, manipulate, and combine data to make it amenable to further analys
es
+ Visualize data with extensive graphics capabilities of R (including ggpl
ot)
+ Use R to run statistical models and hypothesis tests and report results
conforming to standards expected in scientific journals
+ Write reports using the powerful rmarkdown package and its derivatives

### Class outline

```{r}
outline <- tibble(Title = c(
  "Introduction to R: Working environment and data structures",
  "Using packages to enhance data ingestion, munging and reporting",
  "Data visualization for exploration and reporting",
  "Statistical analyses using R",
  "Developing and evaluating models",
  "Reproducible documents for analytic reporting"
)) %>% 
  mutate(`Time (EST)` = '9 - 11 am',
         `Time (UTC)` = '2 - 4 pm',
         Date = format(class_dates, '%d %b')) %>% 
  select(Date, everything())

gt(outline)
```

<p></p>
### The Learning Process

My philosophy is to get students up and running with R for data analysis as quickly as possible. As such, this course is opinionated, in that I make certain choices of what parts of R to teach to make things most accessible and useful.

Students can be successful in this course through following the teaching materials, participating in discussions on Teams, and practice. R is a language in the same way that French or Japanese is a language (you’re just talking to a computer), and so the only way to retain the knowledge gained in this class is to use it. The exercises and tutorials are meant to get you used to using R for different purposes, so please do them diligently.

### Communication

This class will communicate primarily via Teams. We will contain topics for each
week within the chat channel for that week. Common files will be posted in Teams,
but this website will be the prime source of materials. If you have a question, please ask it in the Teams chat so that others can contribute to and learn from the discussion. During the week, I'll
respond to questions and comments at the end of my day (around 4pm EST / 9pm UTC), so if
you don't hear from me sooner, don't panic!!

### Etiquette

The most important thing is to be polite, considerate and empathetic in all communications and discussions. There are different levels of knowledge about R in this class, and so some questions may appear trivial to some but are essential for others. Be kind, and if you can help a classmate, do so with grace and civility. The class learns best if we all help and support each other.
