
---
class: middle,center,inverse

# Reporting modeling results

---

# Tabular results

The **gtsummary** package provides nice tabular summaries of model results, and
can be embeeded in HTML or Word documents (we'll see this in Week 6)

.panelset[
.panel[
.panel-name[Linear regression]
.pull-left[
```{r lmres, eval = F, echo = T}
library(gtsummary)
theme_gtsummary_compact()
tbl_regression(myLinearModel)
```
]
.pull-right[
```{r, eval=T, echo = F, ref.label="lmres"}
```
]
]
.panel[
.panel-name[Logistic regression]
.pull-left[
```{r lrres, eval = F, echo = T}
library(gtsummary)
tbl_regression(myLR, exponentiate = T)
```
]
.pull-right[
```{r, eval=T, echo = F, ref.label="lrres"}
```
]
]
.panel[
.panel-name[Cox regression]
.pull-left[
```{r coxres, eval = F, echo = T}
library(gtsummary)
tbl_regression(survModel, exponentiate = T)
```
]
.pull-right[
```{r, eval=T, echo = F, ref.label="coxres"}
```
]
]
]

---

# A more involved example

```{r, echo=T}
m1 <- lm(body_mass_g ~., data=penguins)
broom::tidy(m1)
```

---

# A quick pipeline to extract factor levels

```{r}
b <- broom::tidy(m1)

# extract feature names
feature_names <- c('\\(Intercept\\)',attr(terms(m1),'term.labels'))
feature_query <- paste(feature_names, collapse='|')
b <- b %>% 
  mutate(features = str_extract(term, feature_query),
         feature_levels = str_remove(term, feature_query)) %>% 
  relocate(features, feature_levels, .after=term)
b
```

---

```{r mod1, include=F}
tbl_regression(m1,
  label = list(bill_length_mm ~ 'Bill length (mm)',
               bill_depth_mm ~ 'Bill depth (mm)',
               flipper_length_mm ~ 'Flipper length (mm)')) %>% 
  add_global_p() %>% 
  bold_p(t = 0.05) %>% 
  bold_labels() %>% 
  italicize_levels()
```

`r chunk_reveal('mod1')`

---


.pull-left[
```{r relevel, echo=T,eval=F}
penguins1 <- penguins %>% 
  mutate(species=fct_relevel(species,'Gentoo')) #<<
m2 <- lm(body_mass_g ~ ., data=penguins1)
tbl_regression(m2,
  label = list(bill_length_mm ~ 'Bill length (mm)',
               bill_depth_mm ~ 'Bill depth (mm)',
               flipper_length_mm ~ 'Flipper length (mm)')) %>% 
  add_global_p() %>% 
  bold_p(t = 0.05) %>% 
  bold_labels() %>% 
  italicize_levels()
```

### Changing reference level for a factor predictor
]
.pull-right[
```{r, ref.label='relevel', echo=F, eval=T, fig.width=2}

```

]

---

### Putting together multiple models

.pull-left[
```{r m1, echo=T, eval=F}
(gt_r1 <- glm(response ~ trt + grade, trial, 
              family = binomial) %>%
  tbl_regression(exponentiate = TRUE))
```
```{r m2, echo=F, eval=F}
(gt_r2 <- coxph(Surv(ttdeath, death) ~ trt + grade,
                trial) %>%
  tbl_regression(exponentiate = TRUE))
```
```{r m3, echo=F, eval=F}
(gt_t1 <- trial[c("trt", "grade")] %>% 
  tbl_summary(missing = "no") %>% 
  add_n() %>%
  modify_header(stat_0 ~ "**n (%)**") %>%
  modify_footnote(stat_0 ~ NA_character_))
```
```{r m4, echo=F, eval=F}
tbl_merge(
  list(gt_t1, gt_r1, gt_r2),
  tab_spanner = c(NA_character_, "**Tumor Response**", "**Time to Death**")
)
```
]
.pull-right[
```{r, eval=T, echo=F, ref.label='m1'}

```
]

---

### Putting together multiple models

.pull-left[
```{r, eval=F, echo=T, ref.label='m2'}
```

]
.pull-right[
```{r, eval=T, echo=F, ref.label='m2'}

```

]

---

### Putting together multiple models

.pull-left[
```{r, eval=F, echo=T, ref.label='m3'}

```

]
.pull-right[
```{r, eval=T, echo=F, ref.label='m3'}

```

]

---

### Putting together multiple models

```{r, eval=T, echo=T, ref.label='m4'}

```

