---
title: "Testing PPTX"
author: "Abhijit Dasgupta"
date: "3/1/2021"
output: 
  powerpoint_presentation: 
    reference_doc: "pptx_droplets.pptx"
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
pacman::p_load(tidyverse, flextable, gtsummary, palmerpenguins)
```

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = F, warning=F, message=F, ft.left=4.5}
tbl_summary(iris) %>% as_flex_table() %>% theme_alafoli()
```

## Slide with Plot

```{r, dpi=150}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species))+
  geom_point()+
  theme_bw()+
  labs(x = 'Sepal Length', y = 'Sepal Width')
```

