---
title: "Week 1 assignment"
author: "Abhijit Dasgupta"
date: "1/25/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading

1. Chapter 2 of Hands on Programming with R ([link](https://rstudio-education.github.io/hopr/basics.html))
1. Parts of Chapters 1 and 2 of the R Cookbook ([link](https://rc2e.com/index.html#the-recipes)) that you find useful. I'd strongly suggest bookmarking section 2.14 ([link](https://rc2e.com/somebasics#recipe-id025))

## Problem 1

Complete the RStudio Primer on Programming Fundamentals available [here](https://rstudio.cloud/learn/primers/1.2)

This primer covers most of what we've covered this week, and a little bit more

## Problem 2

Find the problems (typos) in the following code snippets and correct them so that they run without errors. This exercise will make you aware that the primary reason for R errors is typographical errors. R is case-sensitive and requires the exact specification for each object name. 


```{r, eval=F}
my_variable <- 1:5
my_var1able
```

```{r, eval=F}
data(airquality)
mean(airquality$wind)
```

```{r, eval=F}
table(iris$Sepal_Length)
```

## Problem 3

Write a function that converts miles per gallon to kilometers per liter. The conversion
is 1 mpg = 1.6/3.8 kmpl. The basic format for a function is 

```{r, eval=F}
name <- function(variables) {
  
}
```

Here, you would have one input variable _x_, and it will output one input variable _y_. You need to name your function (please don't keep it as _name_). 

You can then apply this function to the fuel efficiency data in the **mtcars** dataset, which is `mtcars$mpg`. 

## Problem 4

Write a for-loop that generates the natural logarithms (`log`) of the first 20 integers (`1:20`). 

```{r, eval=F}
for(___ in 1:20){
  print(___(____))
}
```

## Problem 5

The `Theoph` data set contains data from an experiment on the pharmacokinetics 
of theophylline. I would like, for each of the 12 subjects in the data, to compute the 
maximum theophylline concentration seen during the experiments.

```{r, eval=F}
data(Theoph)
for(u in unique(Theoph$_____)){
  print(max(Theoph$conc[Theoph$Subject == ____]))
}
```





